---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit README.Rmd and re-knit. -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE, comment = "#>",
  fig.path = "man/figures/README-", out.width = "100%",
  message = FALSE, warning = FALSE
)
```

# HSMap

<!-- badges: start -->
[![R-CMD-check](https://github.com/mmollina/HSMap/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mmollina/HSMap/actions/workflows/R-CMD-check.yaml)
[![Lifecycle: maturing](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://lifecycle.r-lib.org/articles/stages.html#maturing)
<!-- badges: end -->

**HSMap** builds maternal linkage maps from **open-pollinated / unknown-sire diploid
half-sib families**: one or more known dams, many offspring, and unknown fathers.
Transmission along an ordered marker set is modelled with a fast C++ (Rcpp) **maternal
hidden Markov model**. The hidden state is which maternal homolog was transmitted,
transitions encode recombination between adjacent markers, and the unknown paternal
contribution is integrated out of the emissions through a per-marker, **dam-specific
paternal gametic frequency**. When several dams are analysed together, HSMap estimates a
**single shared recombination map** by a joint EM while keeping phase and the paternal
model dam-specific.

Because phase can be locally unresolved (dams are heterozygous at different markers), the
multipoint map is fitted **within resolved phase blocks**, and intervals with no linkage
(recombination fraction at 0.5) are reported as **gaps** rather than as large centimorgan
distances.

> **Scope.** This release covers the open-pollinated / unknown-sire method only.
> Experimental known-sire / full-sib support exists on a separate development branch; it
> is oracle-phase only and **not** part of this release or the accompanying paper.

## Installation

```r
# install.packages("remotes")
remotes::install_github("mmollina/HSMap")
```

## Quick start

The workflow below runs on a small **simulated** example dataset shipped with the package
(no private files required). It covers reading data, two-point analysis, filtering,
grouping, ordering, dam-specific phasing, blockwise multipoint mapping, safe map
reporting, and plotting.

```{r workflow}
library(HSMap)
RcppParallel::setThreadOptions(numThreads = 2)

# 1. Read pedigree + genotype files (simulated example shipped with the package)
ped  <- system.file("extdata", "example_pedigree.csv",  package = "HSMap")
geno <- system.file("extdata", "example_genotypes.csv", package = "HSMap")
dat  <- read_HSMap_data(ped, geno)
dat

# 2. Two-point (pairwise) analysis: recombination fraction + phase LOD for every pair
tpt <- pairwise_rf(dat, threads = 2)

# 3. Filter the two-point table (drop weak / uninformative pairs)
tptf <- tpt_filter(tpt, diagnostic.plot = FALSE)

# 4. Group markers into linkage groups (k = 1 here: the example is one simulated
#    chromosome; use k = <n chromosomes> on real data)
grp <- group_markers(tptf, k = 1, inter = FALSE)

# 5. MDS ordering within each linkage group
ord <- mds_order(grp, tptf, plot_each = FALSE)
lg1 <- ord[[1]]                       # ordered marker IDs of the first linkage group

# 6. Dam-specific phase estimation (one coupling/repulsion configuration per dam)
ph <- phase_from_pairwise(tptf, order = lg1, dam = "all")

# 7. Blockwise multipoint HMM map: the joint EM is fitted WITHIN each resolved phase
#    block, so unresolved phase never forces an imputed map
blocks <- hmm_map_blocks(dat, ph)
blocks$n_blocks

# 8. Safe map positions and lengths: gaps (no-linkage / unresolved) are NA, never a
#    large finite cM distance
bm <- get_block_map(blocks, "haldane")
bm$total_linked_length                # finite linked cM summed across blocks
table(bm$interval_table$status)       # per-interval linked / gap classification
```

```{r plot, fig.alt = "Blockwise linkage map: markers positioned in centiMorgans within resolved phase blocks and map segments.", fig.height = 3.4, fig.width = 7}
# 9. Plot phase blocks and map segments
plot_block_map(blocks, map.function = "haldane")
```

See `vignette("getting-started", package = "HSMap")` for a fuller walk-through, including
the single-dam map and per-offspring haplotype probabilities.

## Citation

If you use HSMap, please cite it (see `citation("HSMap")`). A methods paper is in
preparation.

## License

MIT © the HSMap authors. See [LICENSE](LICENSE).
