---
title: "Getting started with HSMap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with HSMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.2)
```

## Overview

HSMap infers maternal recombination and linkage phase in **open-pollinated diploid
half-sib families**: one or more known dams, many offspring, and **unknown sires**.
Transmission along an ordered marker set is modelled with a maternal-only hidden
Markov model (HMM). The hidden state is the maternal homolog transmitted to an
offspring, transitions encode recombination between adjacent markers, and the
unknown paternal contribution is integrated out of the emissions. When several dams
are analysed together, HSMap estimates a **single shared recombination map** by a
joint EM, while keeping phase and the paternal model dam-specific.

This vignette walks through the workflow on simulated data:

1. simulate a family,
2. two-point (pairwise) analysis,
3. phasing,
4. the multipoint HMM map,
5. the joint multi-dam map,
6. per-offspring haplotype probabilities.

```{r}
library(HSMap)
RcppParallel::setThreadOptions(numThreads = 2)
set.seed(2024)
```

## 1. Simulate a single family

`sim_multi_pop()` simulates families that share one recombination map, with
dam-specific phase and (here) unknown sires drawn under Hardy–Weinberg.

```{r}
Tm <- 40
r_true <- rep(0.04, Tm - 1)                       # ~4 cM (Haldane) per interval
sim1 <- sim_multi_pop(
  T_markers = Tm, n_pops = 1, n_ind_per_pop = 200,
  r_vec = r_true,
  phase_mode = "random", repulsion_rate = 0.3,    # mix of coupling/repulsion
  maternal_geno_mode = "all_het",                 # every marker informative
  paternal_pA_base = 0.4,                          # unknown-sire allele frequency
  error_rate = 0.01, seed = 2024
)
```

Real data is loaded from a pedigree CSV and a genotype CSV with
`read_HSMap_data()`. A small **simulated** open-pollinated example ships with the
package, so the reader runs without any private files:

```{r}
ped  <- system.file("extdata", "example_pedigree.csv",  package = "HSMap")
geno <- system.file("extdata", "example_genotypes.csv", package = "HSMap")
example_dat <- read_HSMap_data(ped, geno)
example_dat
```

An `HSMap.data` object is, at its core, a list of per-family offspring genotype
matrices (`G_list`, coded `0/1/2/NA`) and maternal genotype vectors (`M_list`). For
the rest of this tutorial we wrap the simulated genotypes from Section 1 directly:

```{r}
dat1 <- structure(list(G_list = sim1$G_list, M_list = sim1$M_list),
                  class = "HSMap.data")
markers <- sim1$truth$markers_union
```

## 2. Two-point analysis

`pairwise_rf()` estimates a recombination fraction and a phase LOD for every marker
pair, sharing the recombination fraction across informative families while estimating
a dam-specific paternal gametic frequency and profiling phase per dam.

```{r}
tpt <- pairwise_rf(dat1, threads = 2)
round(tpt$fit$r[1:5, 1:5], 3)
```

For *de novo* analyses you would cluster markers into linkage groups with
`group_markers()` and order each group with `mds_order()`. Here the simulated
markers are already in order, so we use that order directly.

## 3. Phasing

`phase_from_pairwise()` turns the pairwise phase calls into one ordered
coupling/repulsion configuration per dam, via a signed-graph optimisation.

```{r}
ph1 <- phase_from_pairwise(tpt, order = markers)
head(ph1$phase_vec)        # 1 = coupling, 0 = repulsion, per interval
```

## 4. Multipoint HMM map

With phase held fixed, `hmm_map()` fits the maternal HMM by EM. The genotyping
error rate `epsilon` is fixed at the assay's expected value (here 1%).

```{r}
map1 <- hmm_map(dat1, phased = ph1, dam = 1, epsilon = 0.01)
sum(inv_haldane(map1$fit$r))   # total map length in cM (Haldane)
```

```{r, fig.alt = "Linkage map of the single family as a track of marker positions in centiMorgans."}
plot_map_list(map1, map.function = "haldane")
```

## 5. Joint multi-dam map

Now three dams that share one map but have independent phase and unknown,
heterogeneous sires. With `maternal_geno_mode = "HWE"` the dams are heterozygous at
*different* subsets of markers — exactly the case where pooling across families
helps.

```{r}
sim3 <- sim_multi_pop(
  T_markers = Tm, n_pops = 3, n_ind_per_pop = c(120, 100, 80),
  r_vec = r_true,
  phase_mode = "random", repulsion_rate = 0.3,
  maternal_geno_mode = "HWE", maternal_pA = 0.5,
  paternal_pA_base = 0.4, paternal_pA_sd = 0.05,
  error_rate = 0.01, seed = 99
)
dat3 <- structure(list(G_list = sim3$G_list, M_list = sim3$M_list),
                  class = "HSMap.data")
mk3 <- sim3$truth$markers_union

tpt3 <- pairwise_rf(dat3, threads = 2)
ph3  <- phase_from_pairwise(tpt3, order = mk3, dam = "all")   # one entry per dam
```

When dams are heterozygous at different marker subsets, the pairwise phase graph can
break into disconnected components, so some adjacent intervals have an **unresolved**
relative phase (`NA` in `phase_vec`). `hmm_map()` deliberately refuses to run through
unresolved phase; the safe workflow is `hmm_map_blocks()`, which splits the order into
resolved phase blocks and fits the shared joint EM **within each block**:

```{r}
joint <- hmm_map_blocks(dat3, ph3)     # fits the joint map within each resolved block
joint$n_blocks                          # number of resolved phase blocks
# Safe map summary: gaps (no-linkage / unresolved) are NA, never huge cM. Never use
# raw sum(inv_haldane(r)) here, because r may contain a no-linkage boundary.
bm <- get_block_map(joint, "haldane")
bm$total_linked_length                  # finite linked cM summed across blocks
table(bm$interval_table$status)         # linked / gap classification per interval
```

Each block carries a valid ordered marker set with resolved phase for its contributing
dams; intervals between blocks are reported as gaps rather than as large centimorgan
distances. If the whole chromosome is resolved, `hmm_map_blocks()` returns a single
block identical to the ordinary joint `hmm_map()`. A legacy per-dam-then-average
estimator is available for comparison via `hmm_map(..., method = "consensus")` on a
fully resolved order.

## 6. Per-offspring haplotype probabilities

`calc_haploprob()` returns, for each offspring and marker, the posterior probability
of having inherited each maternal homolog. It accepts a fitted map with resolved phase;
here we decode the single-dam map from Section 3:

```{r}
gp <- calc_haploprob(dat1, map1)    # HSMap.gamma for the fitted (resolved) map
dim(gp$gamma)                       # [2 homologs, markers, offspring]
```

These posteriors are the inputs for downstream analyses such as QTL mapping or
crossover counting.

## Session info

```{r}
sessionInfo()
```
