---
title: "Getting started with HSMap"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with HSMap}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(collapse = TRUE, comment = "#>", fig.width = 6, fig.height = 3.2)
```

## Overview

HSMap builds a **maternal linkage map** in a **diploid half-sib /
open-pollinated family**: one known, genotyped dam, many offspring, and
**unknown fathers**. The supported design is exactly this
one-known-maternal-parent case.

Along an ordered marker set, transmission is modelled with a hidden Markov
model. In the production **source-aware** model the hidden state pairs the
transmitted maternal homolog with a paternal *source* state: alongside the
ordinary pollen-population modes (`U0`, `U1`, transmitting paternal allele 0
or 1), the paternal process may occupy a haplotype-**sharing** mode (`H1`,
`H2`) in which the transmitted paternal allele coincides with the allele
carried by one of the dam's own homologs. Such persistent sharing tracts arise
in open-pollinated families when some pollen parents are related to the dam;
modelled explicitly, they are kept out of the maternal recombination map
instead of inflating it.

Two things about the sharing states matter for interpretation:

- they are a **statistical description** of shared-haplotype tracts — not
  father identification, not pedigree reconstruction, and not proof of
  identity by descent;
- the paternal side contributes exactly **two estimated scalars**: the
  probability of entering (`alpha`) and of leaving (`beta`) the sharing mode.
  Setting `alpha = 0` makes the sharing states unreachable and recovers the
  simpler **population-mode** estimator (`hmm_map()`) *exactly*.

This vignette walks the full workflow on simulated data:

1. simulate a family / read data,
2. two-point (pairwise) analysis and filtering,
3. grouping and ordering,
4. phasing,
5. the source-aware chromosome-wide map,
6. population-mode and phase-block-safe fitting,
7. per-offspring maternal inheritance probabilities.

```{r}
library(HSMap)
RcppParallel::setThreadOptions(numThreads = 2)
set.seed(2024)
```

## 1. Simulate a single family

`sim_multi_pop()` simulates families that share one recombination map, with
dam-specific phase and (here) unknown sires drawn under Hardy–Weinberg.

```{r}
Tm <- 40
r_true <- rep(0.04, Tm - 1)                       # ~4 cM (Haldane) per interval
sim1 <- sim_multi_pop(
  T_markers = Tm, n_pops = 1, n_ind_per_pop = 200,
  r_vec = r_true,
  phase_mode = "random", repulsion_rate = 0.3,    # mix of coupling/repulsion
  maternal_geno_mode = "all_het",                 # every marker informative
  paternal_pA_base = 0.4,                          # unknown-sire allele frequency
  error_rate = 0.01, seed = 2024
)
dat1 <- structure(list(G_list = sim1$G_list, M_list = sim1$M_list),
                  class = "HSMap.data")
markers <- sim1$truth$markers_union
```

Real data is loaded from a pedigree CSV and a genotype CSV with
`read_HSMap_data()`. A small **simulated** open-pollinated example ships with
the package, so the reader runs without any private files:

```{r}
ped  <- system.file("extdata", "example_pedigree.csv",  package = "HSMap")
geno <- system.file("extdata", "example_genotypes.csv", package = "HSMap")
example_dat <- read_HSMap_data(ped, geno)
example_dat
```

## 2. Two-point analysis

`pairwise_rf()` estimates a recombination fraction and a phase LOD for every
marker pair, estimating a dam-specific paternal gametic frequency and
profiling phase per dam. `tpt_filter()` screens the two-point table before
grouping and ordering.

```{r}
tpt <- pairwise_rf(dat1, threads = 2)
round(tpt$fit$r[1:5, 1:5], 3)
```

## 3. Grouping and ordering

For *de novo* analyses, cluster markers into linkage groups with
`group_markers()` and order each group with `mds_order()` (both operate on the
filtered two-point object). Here the simulated markers are already in order,
so we use that order directly.

## 4. Phasing

`phase_from_pairwise()` turns the pairwise phase calls into one ordered
coupling/repulsion configuration per dam, via a signed-graph optimisation.

```{r}
ph1 <- phase_from_pairwise(tpt, order = markers)
head(ph1$phase_vec)        # 1 = coupling, 0 = repulsion, per interval
```

## 5. The source-aware chromosome-wide map

With phase held fixed, `hmm_map_source_aware()` fits the production model by
EM: maternal recombination fractions shared by all offspring, plus the two
paternal scalars `alpha` (sharing entry) and `beta` (sharing exit). The
genotyping error rate `epsilon` is fixed at the assay's expected value (here
1%), and the `(alpha, beta)` M-step is exact — it includes the stationary
initial-state term, so each EM iteration has the usual ascent guarantee.

```{r}
sa <- hmm_map_source_aware(dat1, phased = ph1, dam = 1, epsilon = 0.01)
sa
```

The fitted object separates the two processes:

```{r}
sum(sa$dist, na.rm = TRUE)   # maternal map length, cM (Haldane); gaps are NA
c(alpha = sa$alpha, beta = sa$beta)
mean(sa$gammaH)              # posterior sharing-mode occupancy
```

In this simulation the sires are unrelated to the dam, so the estimated
`alpha` is at (or near) its lower bound and the sharing-mode occupancy is
essentially zero — the model collapses to the population-mode estimator, as it
should on data with no sharing. On real open-pollinated data with dam-related
pollen parents, `gammaH` localises the sharing tracts per offspring, and the
map length is protected from their inflation.

Interpretation reminders: `gammaH` is posterior occupancy of a *statistical*
sharing mode; high-occupancy segments flag offspring–region pairs where the
paternal allele stream tracks a dam homolog. They do not identify fathers and
are not, by themselves, evidence of identity by descent.

## 6. Population mode and unresolved phase

`hmm_map()` is the `alpha = 0` special case; it remains the engine for
phase-block-safe fitting. When the pairwise phase graph breaks into
disconnected components (dams heterozygous at different marker subsets), some
adjacent intervals have **unresolved** relative phase, and `hmm_map()` refuses
to run through them. The safe workflow is `hmm_map_blocks()`, which fits
within each resolved phase block:

```{r}
blocks <- hmm_map_blocks(dat1, ph1)
bm1 <- get_block_map(blocks, "haldane")
bm1$total_linked_length             # gap-safe total map length in cM (Haldane)
table(bm1$interval_table$status)    # linked / gap classification per interval
```

Never summarise a map with `sum(inv_haldane(r))` directly: an interval at no
linkage (r near 0.5) would become a huge finite distance. The blockwise
summary reports such intervals as gaps (`NA`), and within-block segments reset
after each gap.

```{r, fig.alt = "Linkage map of the single family as a track of marker positions in centiMorgans."}
map1 <- hmm_map(dat1, phased = ph1, dam = 1, epsilon = 0.01)
plot_map_list(map1, map.function = "haldane")
```

For framework-map work on real data, `diagnose_map_intervals()` classifies
every adjacent interval of a fitted map by multipoint/two-point concordance,
`collapse_framework_blocks()` applies user-directed collapses of locally
discordant blocks, and `try_marker()` / `ripple_map()` provide local order
diagnostics. `get_map()` exports positions under Haldane, Kosambi, or Morgan
map functions.

## 7. Per-offspring maternal inheritance probabilities

`calc_haploprob()` returns, for each offspring and marker, the posterior
probability of having inherited each maternal homolog (the source-aware fit
carries the same information in its `gammaM` element). These posteriors are
the inputs for downstream analyses such as QTL mapping or crossover counting.

```{r}
gp <- calc_haploprob(dat1, map1)    # HSMap.gamma for the fitted (resolved) map
dim(gp$gamma)                       # [2 homologs, markers, offspring]
```

## Extensions (companion paper, forthcoming)

Pooling several dams under one shared map (`hmm_map_joint()`,
`hmm_map_blocks()` on several dams, `test_map_heterogeneity()`) and known-sire
/ full-sib crosses are extensions of this single-family core. They are
included in the software but are documented in a companion paper (forthcoming)
and are not part of the released method's paper.

## Session info

```{r}
sessionInfo()
```
